Antigenic phenotypes and phylogenetic metrics improve forecasts of seasonal influenza A/H3N2 evolution

### 2020-01-03 manuscript/main.tex
Include FRA background and results
Restructures main text to introduce FRA assays early on as a more recent
alternative to HI assays, describe the limited availability of FRA data for
training/validation/testing along HI models, include FRA results in the final
results section, and reduce the description of FRAs as future directions in the
discussion. This commit also include an additional citation for the
glycosylation motifs in modern H3N2s (Chambers et al. 2015) and a citation for
the FRAs (Okuno et al. 1990).

Seasonal influenza virus infects 5--15\% of the global population every year causing an estimated 250,000 to 500,000 deaths annually \cite{flufactsheet}.
Vaccination remains the most effective public health response available.
However, frequent viral mutation results in viruses that escape previously acquired human immunity.
The World Health Organization (WHO) selects vaccine viruses to match circulating viruses, but because the process of vaccine development and distribution requires several months to complete, accurate vaccine strain selection requires a prediction of which viruses will predominate approximately one year after vaccine viruses are selected.
Current vaccine predictions focus on the hemagglutinin (HA) protein, which acts as the primary target of human immunity.
The hemagglutination inhibition (HI) assay \cite{hirst1943studies} is used to measure the degree of cross-reactivity between pairs of circulating viruses.
HI assays are fundamental for vaccine strain selection, but they are laborious and low-throughput compared to genome sequencing \cite{Wood:2012ii}.
As a result, researchers have developed computational methods to predict influenza fitness from sequence data alone \cite{Luksza:2014hj,Steinbruck:2014kq,Neher:2014eu}.

Despite the promise of these sequence-only models, they explicitly omit experimental measurements of antigenic or functional phenotypes.
Recent developments in computational methods and influenza virology have made it feasible to integrate these important metrics of influenza fitness into a single predictive model.
For example, phenotypic measurements of antigenic drift are now accessible through phylogenetic models \cite{Neher:2016hy} and functional phenotypes for HA are available from deep mutational scanning experiments \cite{Lee2018}.
We describe an approach to integrate previously disparate sequence-only models of influenza evolution with high-quality experimental measurements of antigenic drift and functional constraint.

The influenza community has long recognized the importance of incorporating HI phenotypes and other experimental measurements of viral phenotypes with existing forecasting methods into an extensible, open source framework that can be used by professional virologists in their vaccine design process \cite{Gandon:2016gz,Morris:2017ea,Lassig:2017hr}.
Although several distinct efforts have made progress in using HI phenotypes to evaluate the evolution of seasonal influenza \cite{Steinbruck:2014kq,Neher:2016hy}, these methods stop short of developing a complete forecasting framework wherein the evolutionary contribution of HI phenotypes can be compared and contrasted with new and existing fitness metrics.
Here, we provide the first such open source long-term forecasting framework for seasonal influenza.
With this framework, we show that HI phenotypes enable more accurate long-term forecasts of A/H3N2 populations compared to previous metrics based on epitope mutations alone.
However, we also find that composite models based on phenotypic measures of antigenic drift and genotypic measures of functional constraint consistently outperform any fitness models based on individual genotypic or phenotypic metrics.

@@

Next, we trained and validated models for individual fitness predictors using 25 years of natural H3N2 populations spanning from October 1, 1990 to October 1, 2015.
We held out samples collected between October 1, 2015 and October 1, 2019 for model testing (Supplemental Fig.~\ref{sup_fig:cross_validation_for_natural_populations}).
In addition to the sequence-only models we tested on simulated populations, we also fit models for our new fitness metrics based on experimental phenotypes including HI cross-immunity and DMS mutational effects.
We hypothesized that both HI and DMS metrics would be assigned positive coefficients, as they estimate increased antigenic drift and beneficial mutations, respectively.
As antigenic drift is generally considered to be the primary evolutionary pressure on natural H3N2 populations \cite{Smith:2004jc,Bedford:2014bf,Luksza:2014hj}, we expected that epitope and HI cross-immunity would be individually more predictive than mutational load or DMS mutational effects.
Previous research \cite{Neher:2016hy} and our simulation results also led us to expect that LBI and delta frequency would outperform other individual mechanistic metrics.

@@

The stability of the coefficients for the metrics in the two-metric models suggested that these metrics represented complementary components of viral fitness.
In contrast, the three-metric model strongly preferred the HI cross-immunity and mutational load metrics over LBI for the majority of the validation period, producing an average LBI coefficient of 0.05 $\pm$ 0.50.
In the last two validation timepoints, this composite model converged to a nearly LBI-only model with the coefficients for the other two metrics converging to near-zero values.
These last two timepoints span the period when an HA1:160T substitution swept through the global population.
This substitution introduced a new glycosylation motif and dramatically reduced the efficiency of HI assays \cite{Zost2017}.
These results reinforce the historical importance of HI assays in measuring viral fitness, while also indicating that these assays may no longer be as predictive as sequence-only metrics.

@@

\subsection*{Forecasts predict the rise of A1b/131K and A1b/135K sub-clades in September 2020}

To enable real-time forecasts, we integrated our forecasting framework into our existing open source pathogen surveillance application, Nextstrain \cite{Hadfield2018}, and used our best model to forecast H3N2 populations in September 2020.
Prior to finalizing our model coefficients for use in Nextstrain, we tested whether our three best composite models could be improved by learning coefficients per timepoint from the test data.
None of the three models performed better across the test data than the original best model, indicating that the coefficients for our selected best model would be most accurate for live forecasts (Supplemental Fig.~\ref{sup_fig:updated_model_accuracy_and_coefficients_for_natural_populations_across_test_data}).
With the original coefficients, our combined HI cross-immunity and mutational load model favored the growth of the currently circulating clade A1b/197R, the maintenance of A1b/137F, and the decline of 3c3.A (Fig.~\ref{fig:nextstrain_forecasts}).
To aid with identification of potential vaccine candidates, we annotated samples in the phylogeny by their estimated distance to the future based on our best model (Fig.~\ref{fig:nextstrain_distance_to_future}).

@@

Several clear opportunities to improve forecasts still remain.
Integration of more recent experimental data may improve estimates of antigenic drift.
Influenza collaborating centers now routinely measure antigenic drift by focus reduction assays (FRAs), as an alternative to hemagglutination inhibition (HI) assays that is less affected by glycosylation motifs that alter binding efficacy \cite{Zost2017}.
While there are not enough historical FRA measurements to extensively validate and test a FRA cross-immunity metric, it may be possible to reuse coefficients from the HI cross-immunity models with modern FRA data.
Further experiments will be required to fully understand how HI and FRA measurements of the same pairs of strains correlate with each other.
In addition to these FRA data based on ferret antisera, recent high-throughput antigenic escape assays with human sera promise to improve existing definitions of epitope sites \cite{Lee2019}.
These assays reveal the specific sites and residues that confer antigenic escape from polyclonal sera obtained from individual humans.
A sufficiently broad geographic and temporal sample of human sera with these assays could reveal consistent patterns of the immune landscape H3N2 strains must navigate to be globally successful.

Finally, forecasting models need to account for the geographic distribution of viruses.
Globally successful clades of influenza H3N2 most commonly emerge from Southern Asia, while viruses emerging from Africa, North America, and Europe are less likely to fix globally \cite{Russell:2008ke,Rambaut:2008ew,Bedford:2015fj}.
Explicitly accounting for these source/sink dynamics would allow models to weight forecasts based on both viral fitness and transmission.
For example, a geographically-aware implementation of LBI would most likely have not incorrectly predicted the success of the predominantly North American and European reassortant clade A2/re in 2019.

>>>>

Seasonal influenza virus infects 5--15\% of the global population every year causing an estimated 250,000 to 500,000 deaths annually with the majority of infections caused by influenza A/H3N2 \cite{flufactsheet}.
Vaccination remains the most effective public health response available.
However, frequent viral mutation results in viruses that escape previously acquired human immunity.
The World Health Organization (WHO) selects vaccine viruses to match circulating viruses, but because the process of vaccine development and distribution requires several months to complete, accurate vaccine strain selection requires a prediction of which viruses will predominate approximately one year after vaccine viruses are selected.
Current vaccine predictions focus on the hemagglutinin (HA) protein, which acts as the primary target of human immunity.
Until recently, the hemagglutination inhibition (HI) assay has been the primary experimental measure of antigenic cross-reactivity between pairs of circulating viruses \cite{hirst1943studies}.
Most modern H3N2 strains carry a glycosylation motif that reduces their binding efficiency in HI assays \cite{Chambers:2015jt,Zost2017}.
Correspondingly, many influenza collaborating centers now also measure cross-reactivity by focus reduction assays (FRAs) \cite{Okuno:1990uu}.
Together, these two assays are the gold standard for vaccine strain selection, but they are laborious and low-throughput compared to genome sequencing \cite{Wood:2012ii}.
As a result, researchers have developed computational methods to predict influenza fitness from sequence data alone \cite{Luksza:2014hj,Steinbruck:2014kq,Neher:2014eu}.

Despite the promise of these sequence-only models, they explicitly omit experimental measurements of antigenic or functional phenotypes.
Recent developments in computational methods and influenza virology have made it feasible to integrate these important metrics of influenza fitness into a single predictive model.
For example, phenotypic measurements of antigenic drift are now accessible through phylogenetic models \cite{Neher:2016hy} and functional phenotypes for HA are available from deep mutational scanning (DMS) experiments \cite{Lee2018}.
We describe an approach to integrate previously disparate sequence-only models of influenza evolution with high-quality experimental measurements of antigenic drift and functional constraint.

The influenza community has long recognized the importance of incorporating HI phenotypes and other experimental measurements of viral phenotypes with existing forecasting methods into an extensible, open source framework that can be used by professional virologists in their vaccine design process \cite{Gandon:2016gz,Morris:2017ea,Lassig:2017hr}.
Although several distinct efforts have made progress in using HI phenotypes to evaluate the evolution of seasonal influenza \cite{Steinbruck:2014kq,Neher:2016hy}, these methods stop short of developing a complete forecasting framework wherein the evolutionary contribution of HI phenotypes can be compared and contrasted with new and existing fitness metrics.
Here, we provide the first such open source long-term forecasting framework for seasonal influenza.
With this framework, we show that HI phenotypes enable more accurate long-term forecasts of H3N2 populations compared to previous metrics based on epitope mutations alone.
However, we also find that composite models based on phenotypic measures of antigenic drift and genotypic measures of functional constraint consistently outperform any fitness models based on individual genotypic or phenotypic metrics.

@@

Next, we trained and validated models for individual fitness predictors using 25 years of natural H3N2 populations spanning from October 1, 1990 to October 1, 2015.
We held out samples collected between October 1, 2015 and October 1, 2019 for model testing (Supplemental Fig.~\ref{sup_fig:cross_validation_for_natural_populations}).
In addition to the sequence-only models we tested on simulated populations, we also fit models for our new fitness metrics based on experimental phenotypes including HI cross-immunity and DMS mutational effects.
We hypothesized that both HI and DMS metrics would be assigned positive coefficients, as they estimate increased antigenic drift and beneficial mutations, respectively.
As antigenic drift is generally considered to be the primary evolutionary pressure on natural H3N2 populations \cite{Smith:2004jc,Bedford:2014bf,Luksza:2014hj}, we expected that epitope and HI cross-immunity would be individually more predictive than mutational load or DMS mutational effects.
Previous research \cite{Neher:2016hy} and our simulation results also led us to expect that LBI and delta frequency would outperform other individual mechanistic metrics.
As the earliest measurements from focus reduction assays (FRAs) date back to 2012, we could not train, validate, and test FRA cross-immunity models in parallel with the HI cross-immunity models.

@@

The stability of the coefficients for the metrics in the two-metric models suggested that these metrics represented complementary components of viral fitness.
In contrast, the three-metric model strongly preferred the HI cross-immunity and mutational load metrics over LBI for the majority of the validation period, producing an average LBI coefficient of 0.05 $\pm$ 0.50.
In the last two validation timepoints, this composite model converged to a nearly LBI-only model with the coefficients for the other two metrics converging to near-zero values.
These results suggest that HI assays and LBI may measure similar fitness effects.
However, the dominance of LBI later in this three-metric model also suggests that HI assays may no longer be as predictive as they have been historically.

@@

\subsection*{Historically-trained models enable real-time, actionable forecasts}

To enable real-time forecasts, we integrated our forecasting framework into our existing open source pathogen surveillance application, Nextstrain \cite{Hadfield2018}.
Prior to finalizing our model coefficients for use in Nextstrain, we tested whether our three best composite models could be improved by learning new coefficients per timepoint from the test data.
In addition to three composite models described above, we evaluated a composite of FRA cross-immunity and mutational load.
Since the earliest FRA data are from 2012, we anticipated that there were enough measurements to fit a model across the test data time interval.
If modern H3N2 strains continue to perform poorly in HI assays, the FRA-based assay will be critical for future forecasting efforts.

None of the three original composite models performed better across the test data than the original best model (Supplemental Fig.~\ref{sup_fig:updated_model_accuracy_and_coefficients_for_natural_populations_across_test_data}).
These results confirmed that the coefficients for our selected best model would be most accurate for live forecasts.
Interestingly, the FRA cross-immunity metric received a consistently positive coefficient in its composite with mutational load and this composite ranked among the best models for the test data.
These results indicate that this FRA-based composite could eventually replace the HI-based composite with mutational load.

After confirming the coefficients for our best model of HI cross-immunity and mutational load, we inspected forecasts of H3N2 clades from January 2020 to 2021.
Of the three major currently circulating clades, our model predicted the growth of A1b/197R, the maintenance of A1b/137F, and the decline of 3c3.A (Fig.~\ref{fig:nextstrain_forecasts}).
To aid with identification of potential vaccine candidates for the next season, we annotated samples in the phylogeny by their estimated distance to the future based on our best model (Fig.~\ref{fig:nextstrain_distance_to_future}).

@@

Several clear opportunities to improve forecasts still remain.
Integration of more recent experimental data may improve estimates of antigenic drift.
The strong performance of a composite FRA cross-immunity and mutational load model with recent H3N2 populations indicates that FRAs will continue to be important for future forecasting efforts.
In addition to these FRA data based on ferret antisera, recent high-throughput antigenic escape assays with human sera promise to improve existing definitions of epitope sites \cite{Lee2019}.
These assays reveal the specific sites and residues that confer antigenic escape from polyclonal sera obtained from individual humans.
A sufficiently broad geographic and temporal sample of human sera with these assays could reveal consistent patterns of the immune landscape H3N2 strains must navigate to be globally successful.
Finally, forecasting models need to account for the geographic distribution of viruses.
Globally successful clades of influenza H3N2 most commonly emerge from Southern Asia, while viruses emerging from Africa, North America, and Europe are less likely to fix globally \cite{Russell:2008ke,Rambaut:2008ew,Bedford:2015fj}.
Explicitly accounting for these source/sink dynamics would allow models to weight forecasts based on both viral fitness and transmission.

### 2020-01-03 manuscript/supplement.tex
Include FRA background and results
Restructures main text to introduce FRA assays early on as a more recent
alternative to HI assays, describe the limited availability of FRA data for
training/validation/testing along HI models, include FRA results in the final
results section, and reduce the description of FRAs as future directions in the
discussion. This commit also include an additional citation for the
glycosylation motifs in modern H3N2s (Chambers et al. 2015) and a citation for
the FRAs (Okuno et al. 1990).

\begin{figure*}[ht]
  \begin{center}
  \includegraphics[width=\textwidth]{figures/models-natural-populations-composite-with-updated-coefficients-across-test-data.png}
  \caption{
    Model coefficients and distances to the future for best composite models fit to recent data from natural populations as in Fig.~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_simulated_populations_controls}.
    A) Coefficients and B) distances are shown per test timepoint (N=8).
    In contrast to the results for these models based on fixed coefficients from training/validation, these coefficients were learned for each six-year window prior to the corresponding test timepoint.
    The corresponding distances reflect the model's performance with updated coefficients on what is effectively new validation data.
    The naive model's distance to the future was 6.82 $\pm$ 1.74 AAs for these timepoints.
  }
  \label{sup_fig:updated_model_accuracy_and_coefficients_for_natural_populations_across_test_data}
  \end{center}
\end{figure*}

>>>>

\begin{figure*}[ht]
  \begin{center}
  \includegraphics[width=\textwidth]{figures/models-natural-populations-composite-with-updated-coefficients-across-test-data.png}
  \caption{
    Model coefficients and distances to the future for best composite models and a FRA-based composite fit to recent data from natural populations as in Fig.~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_simulated_populations_controls}.
    A) Coefficients and B) distances are shown per test timepoint (N=8).
    In contrast to the results for these models based on fixed coefficients from training/validation, these coefficients were learned for each six-year window prior to the corresponding test timepoint.
    The corresponding distances reflect the model's performance with updated coefficients on what is effectively new validation data.
    The naive model's distance to the future was 6.82 $\pm$ 1.74 AAs for these timepoints.
  }
  \label{sup_fig:updated_model_accuracy_and_coefficients_for_natural_populations_across_test_data}
  \end{center}
\end{figure*}
